Online methods

### 2013-09-13 docs/manuscript/manuscript.tex
The main text and methods have been updated.

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects (Supplementary Figure \ref{fig:ld_mean}) \cite{Weir2008, Hemani2013}, and for higher power terms of LD $r$ the sampling variance increases (Methods, Supplementary Figure \ref{fig:ld_sd}). Therefore these results are encouraging with regards to the detection and replication of epistasis.

@@

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold. How does this compare to the number of traits influenced by additive effects? The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL). Therefore it can be argued that the number of instances of detectable epistasis is substantial.

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that on average they explain 1.73\% of the phenotypic variance of each of the 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance. There are several caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_mean}
	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
	\label{fig:ld_mean}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_sd}
	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As the power term of $r$ increases the sampling variance also increases. Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
	\label{fig:ld_sd}
\end{figure}
\clearpage

>>>>

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).

It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice for several reasons. For example, LD between causal variants and observed markers plays an important role. Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication. Indeed, in addition to heterogeneity between populations, sampling variance of LD $r$ alone leads to over three fold decrease in observed variance for the most extreme case of $D \times D$ effects when compared to the average decrease for additive variance on average (Methods, Supplementary Figure \ref{fig:ld_reduction}). Therefore these results are encouraging with regards to the detection and replication of epistasis.

@@

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking published results from a standard eQTL study \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) where the minimum additive variance was $2.1\%$, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain a total of 0.22\% of phenotypic variance, approximately ten times lower than the additive variance. There are several caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_mean}
	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
	\label{fig:ld_sampling_mean}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_sd}
	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As the power term of $r$ increases the sampling variance also increases. Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
	\label{fig:ld_sampling_sd}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_reduction}
	\caption{\textbf{Reduction in LD as estimated in replication data after ascertaining for high LD in discovery data} 100,000 ``unobserved'' causal variants (CVs) were tested for LD against a panel of 528,509 ``observed'' discovery markers (DMs). DM/CV pairs with LD $r^2 > 0.9$ were then tested in an independent sample. Simulation results of the proportional decrease between discovery and replication datasets in LD ($y$-axis) of $\hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ ($x$-axis) are shown, where $\hat{r}^{x}_{D}$ and $\hat{r}^{x}_{R}$ are the sample LD measurements in the discovery and replication datasets, respectively. The average proportional decrease in the replication $\hat{r}^{x}_{R}$ was $2.8\%, 5.3\%, 7.4\%$ and $9.2\%$ for $x=2,4,6$ and $8$, respectively.}
	\label{fig:ld_reduction}
\end{figure}
\clearpage

### 2013-09-13 docs/manuscript/methods.tex
The main text and methods have been updated.

Simulation 1 shows that sampling variance of $r^x$ increases as $x$ increases. Detection of epistatis is highly dependent upon high $\hat{r}$. Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$. Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication. These results demonstrate that assuming all else is equal, sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.

>>>>

Simulation 1 shows that sampling variance of $r^x$ increases as $x$ increases. Detection of epistatis is highly dependent upon high $\hat{r}$. Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$. Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication. These results demonstrate that if all else is equal, the impact of sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
